---
title: "Enriching and plotting a networkGen network"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Enriching and plotting a networkGen network}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

`networkPlot` takes a network built by `networkGen` and produces two things:
its **pathway enrichment** and its **interactive HTML rendering** (plus
kinase-by-pathway heatmaps). This vignette covers the inputs, the
enrichment call, reading the pathway output, and every knob on the HTML.

```{r setup}
library(networkPlot)
```

## 1. Inputs

Everything starts from a `networkGen` result -- its `nodes` and `edges`
data frames. `nodes` carries `Protein`, `type` (`Kinase`, `Sensitivity`,
`Protein-Kinase`, ..., or `Hidden` for the Steiner connector nodes PCSF
added), `prize`, `LogFC`, `degree`. `edges` carries `from`, `to`, `weight`
(how many of the randomized PCSF runs kept the edge) and `cost` (the
reference PPI's interaction cost -- lower = a stronger interaction).

```{r toy}
nodes <- data.frame(
  Protein = c("EGFR", "AKT1", "MTOR", "RPS6KB1", "MAPK1", "BRAF", "STEINER1"),
  type    = c("Protein-Kinase", "Kinase", "Kinase", "Kinase", "Kinase", "Kinase", "Hidden"),
  prize   = c(0.95, 0.9, 0.85, 0.7, 0.8, 0.75, 2),
  LogFC   = c(2.4, 1.8, -1.2, -2.0, 1.1, 0.6, 0),
  degree  = c(6, 5, 4, 2, 3, 2, 4)
)
edges <- data.frame(
  from = c("EGFR", "EGFR", "AKT1", "MTOR", "MAPK1", "STEINER1"),
  to   = c("AKT1", "MAPK1", "MTOR", "RPS6KB1", "STEINER1", "BRAF"),
  weight = c(8, 7, 8, 5, 3, 6),
  cost   = c(0.08, 0.12, 0.05, 0.30, 0.45, 0.10)
)
```

For enrichment you also need the Reactome reference tables. A snapshot
ships with the package -- `reactome_refs()` loads it. Refresh it when
Reactome cuts a new release with `download_reactome_refs(dir)`; it is never
fetched automatically.

## 2. Enrichment

`enrich_network()` runs one Enrichr query (via the `enrichR` package) over
**every node of the network except the Steiner connectors** (`type ==
"Hidden"`), keeps pathways whose gene overlap is at least `min_hits`, and
returns the pathway table plus a `pathway` column joined onto `nodes`.

**Only Reactome is wired end to end today.** The reference database is a
swap point: `databases` is passed straight to Enrichr, which ships many
libraries (KEGG, WikiPathways, GO, MSigDB Hallmark, ...). Anything
Reactome-specific -- collapsing redundant terms by shared parent, dropping
over-general top-level terms, filtering WikiPathways by ontology tag --
lives in the `postprocess` argument (`reactome_postprocess()` by default;
pass `NULL` to skip it). Trying the other Enrichr libraries is a planned
next step.

```{r enrich}
res <- enrich_network(nodes, databases = "Reactome_Pathways_2024", min_hits = 2)
```

`res$short` is one pathway per gene set; the full table can be long, so
just a peek here -- see the next section for what each column means:

```{r enrich-table, echo = FALSE}
knitr::kable(head(res$short, 2), digits = 3, row.names = FALSE, caption = "Top enriched pathways (first 2 rows)")
```

Pass `out_dir` and `label` (e.g. `label = "DrugA_vs_DMSO_spec0.7"`) to also
write `pathways_<label>_all.csv` (every candidate) and `pathways_<label>.csv`
(the *provisional* one-pathway-per-gene-set pick shown above) to disk --
omitted here so this vignette doesn't write files. Once every comparison
for a `spec_cutoff` is done, `reconcile_pathways_dir(res_dir, spec_cutoff)`
re-reads the `_all` files and rewrites each short table so the pathway that
recurs across the most comparisons wins -- that way the same biology sits
under the same label in every comparison, and the combined heatmaps line
up. Re-running with `refresh = FALSE` (the default) skips the Enrichr call
and reads the existing `pathways_<label>.csv` back instead.

## 3. Reading the pathway output

Each row of the pathway table is one enriched pathway: `Pathway` /
`Group_Pw` (the name, `Group_Pw` prefixed with its Reactome top group),
`Genes` (the network genes in it, `;`-joined), `n_hits` / `pathway_size` /
`overlap`, `Adjusted.P.value.mean`, `Combined.Score.mean`.

`plot_pathway_heatmaps(comparisons, out_dir)` (or the folder wrapper
`plot_pathway_heatmaps_dir()`) draws, per `spec_cutoff`: one pathway x
kinase `LogFC` heatmap per comparison, a heatmap of the pathways common to
*all* comparisons, and -- for two comparisons -- a "unique to each"
heatmap, or for three-plus, one pairwise-overlap heatmap per pair.

## 4. The interactive HTML

`plot_network()` returns a `visNetwork` htmlwidget -- printing it (or, as
here, just letting it be the chunk's last value) embeds the live,
interactive network right in this page. Plotting `res$nodes` (from the
enrichment step above, so it carries the `pathway` column) turns on the
pathway multi-select dropdown -- try it:

```{r plot}
w <- plot_network(
  res$nodes, edges, title = "DrugA vs DMSO",
  theme = networkplot_theme(edge_width = list(column = "cost", invert = TRUE))
)
w
```

```{r save, eval = FALSE}
save_network_html(w, "results/DrugA/DrugA_vs_DMSO.html")
```

### Where the theme lives

Every visual choice is a field on the list `networkplot_theme()` returns,
passed as `theme =`. Calling it with no arguments gives the standard look;
override a field by name to change just that one thing.

```{r theme}
th <- networkplot_theme(
  shapes     = c(Kinase = "diamond"),            # merged over the defaults
  edge_width = list(column = "cost", invert = TRUE)  # thicker edge = stronger interaction
)
names(th)
```

Tunable HTML features and their theme fields:

| Feature | Field(s) |
|---|---|
| node shape per `type` | `shapes` (merged over the default map), `shape_default` |
| node fill = diverging `LogFC` scale | `logfc_palette`, `logfc_limit_fn` |
| high-degree node highlight | `highlight_degree`, `highlight_border`, `border_default`, `highlight_border_width`, `border_width_default` |
| edge colour | `edge_colour` |
| **edge boldness** | `edge_width` -- `NULL` for constant, or `list(column, scale, invert)`; use `column = "cost", invert = TRUE` for interaction strength, or `column = "weight"` for run-frequency robustness |
| fonts | `node_font_size`, `legend_font_size` |
| layout | `layout`, `seed` |

The pathway multi-select dropdown and the node-search box are always on;
the dropdown appears whenever `nodes` has a `pathway` column and
`colour_by = "pathway"` (the default).

## 5. End to end, in one call

Enriching and plotting is almost always done together, so
`enrich_and_plot_network()` wraps both steps -- call it once and get both
results back (same result as sections 2 and 4 combined, so not re-run here):

```{r e2e, eval = FALSE}
out <- enrich_and_plot_network(
  nodes, edges, title = "DrugA vs DMSO",
  theme = networkplot_theme(edge_width = list(column = "cost", invert = TRUE))
)
save_network_html(out$widget, "results/DrugA/DrugA_vs_DMSO.html")
```

Call `enrich_network()` and `plot_network()` separately instead of the
wrapper when you need `reconcile_pathways()` across several comparisons
before plotting any of them, or already have an enrichment result from
elsewhere.
